---
title: "Beatles Data Story"
output:
  html_document:
    csl: style.cls
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The Bealtes have been a household name for decades. 
They are regarded by many as one of the great rock bands of all time, but what was it that made them so popular? 
Perhaps some exploratory data analysis might provide an answer.


Let's load our necessary, packages first.
```{r}
library(tidyverse)
library(tm)
library(wordcloud)
library(tidytext)
library(ggplot2)
library(syuzhet)
library(plotly)
library(data.table)
```
Then we shall load our processed lyrics. 
```{r}
load("../output/processed_lyrics.RData")

artists <- read.csv("../data/artists.csv")
```


For this analysis, I'm focusing on the beatles within the larger subgroup of rock artists formed between 1960 and 1970 (inclusive).  
I'm splitting the data into beatles and non-beatles for comparison and contrast.

```{r}

artists_60_70s<- artists %>% 
  filter(Formed %in% c(1960:1970)) %>% 
  select(Artist)

beatles_lyrics <- dt_lyrics %>% 
  filter(artist =="beatles" & genre == "Rock")

other_lyrics <- dt_lyrics %>% 
  filter(artist %in% artists_60_70s$Artist & genre == "Rock" & artist != "beatles") 
```

When comparing the number of stemmed words per song for each group, it seems that the Beatles have a lower average count (51) than similar rock bands (63). 

```{r}
# the number of stemmed words in each song by group
beatles.words_per_song <- sort(sapply(strsplit(beatles_lyrics$stemmedwords, " "), length))
other.words_per_song <- sort(sapply(strsplit(other_lyrics$stemmedwords, " "), length))


# mean, median, and spread
summary(beatles.words_per_song)
summary(other.words_per_song)



# resulting plot
hist(beatles.words_per_song, main = "Beatles Song lengths (by stemmed words)", xlab = "Number of words in a song", col = "blue", breaks = 100)
hist(other.words_per_song, main = "Other Song lengths (by stemmed words)", xlab = "Number of words in a song", col = "red", breaks = 100)
boxplot(beatles.words_per_song, other.words_per_song, horizontal = T, names = c("beatles", "other"), col = c("blue", "red"), xlab= "Number of words in a song")
```

But what if the other bands' distribution is skewed by outliers? Even if we remove the higher outliers, the Beatles still have a lower average number of stemmed words. This may imply that Beatles' songs have on average shorter lyrics.

```{r}


# finding outliers for each group
outliers1 <- boxplot(beatles.words_per_song, plot = F)$out
outliers2 <- boxplot(other.words_per_song, plot = F)$out

# remove outliers 

beatles.words_per_song2 <- beatles.words_per_song[-which(beatles.words_per_song %in% outliers1)]
other.words_per_song2 <-  other.words_per_song[-which(other.words_per_song %in% outliers2)]
# mean, median, and spread
summary(beatles.words_per_song2)
summary(other.words_per_song2)

# resulting plot
hist(beatles.words_per_song, main = "Beatles Song lengths (by stemmed words) ~ deleted outliers", xlab = "Number of words in a song", col = "blue", breaks = 100)
hist(other.words_per_song, main = "Other Song lengths (by stemmed words) ~ deleted outliers", xlab = "Number of words in a song", col = "red", breaks = 100)
boxplot(beatles.words_per_song2, other.words_per_song2, horizontal = T, names = c("Beatles", "Other"), xlab = "Number of words per song", col = c("blue", "red"))
```


Next, I'll convert the given stemmed words into a corpus and then a term document matrix, since the data was already processed and cleaned.

```{r}
# Created a function to streamline conversion to term document matrix

stem_to_corpus <- function(x){
  source <- VectorSource(x)
  corpus <- VCorpus(source)
  return (corpus)
}

# respective corpi
beatles.corpus <- stem_to_corpus(beatles_lyrics$stemmedwords)
other.corpus <- stem_to_corpus(other_lyrics$stemmedwords)
# respective tdm
beatles.tdm <- TermDocumentMatrix(beatles.corpus)
other.tdm <- TermDocumentMatrix(other.corpus)

#beatles.tdm <- removeSparseTerms(beatles.tdm, 0.99) # remove lower frequency terms
#other.tdm <- removeSparseTerms(other.tdm, 0.99) # remove lower frequency terms

# Conveting TDMs into dataframes of most frequent terms

# beatles
m.beatles <- as.matrix(beatles.tdm)
f.b <- sort(rowSums(m.beatles), decreasing=T)
beatles.word_freq<- data.frame(word= names(f.b), freq=f.b)

# other
m.other <- as.matrix(other.tdm)
f.other <- sort(rowSums(m.other), decreasing=T)
other.word_freq<- data.frame(word= names(f.other), freq=f.other)
```


Here are the respective word clouds.

```{r}
# wordcloud for beatles top 100 words
set.seed(1)
wordcloud(words = beatles.word_freq$word, freq =beatles.word_freq$freq, max.words = 100, random.order=FALSE,
          colors=brewer.pal(8, "Dark2"))

# wordcloud for other top 100 words
set.seed(1)
wordcloud(words = other.word_freq$word, freq =other.word_freq$freq, max.words = 100, random.order=FALSE,  
          colors=brewer.pal(8, "Dark2"))

```

Without even looking at any quantitative summaries, it's clear that "love" is the most dominant word in both Beatles and Other rock groups.



```{r}
# top 20 terms
head(beatles.word_freq, 20)
head(other.word_freq, 20)
```


Based on the top twenty words in the respective dataframes, it seems that there is little difference between the most used words from Beatles lyrics and those of other rock groups of the sixties/seventies. 


The words for the most part seem synonymous with similar frequencies; though there is an exception for "cry" (possibly invoking a negative sentiment) which is the 10th most used word in the Beatles lyrics and 20th in other bands.

The relative frequency in the bar plots, table, boxplots and piecharts indicate there is some difference in the distributions of the Beatles words. For example, in terms of frequency, Love is used 976 times in Beatles lyrics and 6432 times in other bands' lyrics; however when looking at their relative frequencies, we can see that the Beatles use love nearly twice as much as Other (.053 vs. .024).

```{r}

# total number of words for each group
n1<-sum(beatles.word_freq$freq)
n2<-sum(other.word_freq$freq) 

# relative frequencies  for each group
beatles.rf20 <- round(head(beatles.word_freq$freq, 20)/n1, 3)
other.rf20 <- round(head(other.word_freq$freq, 20)/n2, 3)

# relative frequency table
data.table(cbind("Beatle words" = as.character(head(beatles.word_freq$word, 20)), 
                 "freq" = head(beatles.word_freq$freq, 20),
                 "r-freq" = head(beatles.word_freq$freq, 20)/n1))
data.table(cbind("Other words" = as.character(head(other.word_freq$word, 20)),
                 "freq" = head(other.word_freq$freq, 20),
                 "r-freq" = head(other.word_freq$freq, 20)/n2))
 
# barplots with 
barplot(beatles.rf20, las = 1,names.arg = head(beatles.word_freq$word, 20), col = rainbow(20), horiz = T, main = "Relative Frequency of top 20 words in Beatles songs")
barplot(other.rf20, las = 2,names.arg = head(other.word_freq$word, 20), col = rainbow(20),horiz = T, main = "Relative Frequency of top 20 words in Other songs")

```

Perhaps a more direct comparison will help.

```{r}

# save top 50 words as character strings

a<- as.character(head(beatles.word_freq$word, 20))  
b <- as.character(head(other.word_freq$word, 20)) 

# intersection of common popular words for beatles and other
v<- a[a %in% b] # 14 words

# return in each group the word 
o<-other.word_freq %>% 
  filter(word %in% v) 

b<-beatles.word_freq %>% 
  filter( word %in% v) 

# merge frequencies by words
top14 <- merge(b, o, by="word")  
top14 <- top14 %>% 
  rename(
    beatles = freq.x,
    other = freq.y
  )
top14

# relative frequency
#barplot(c(head(top14$beatles,10) , head(top14$other,10)) , las = 2,names.arg = c(
#  as.character(head(top14$word,10)),
#  as.character(head(top14$word,10))),col = c("red","blue"), horiz = T, 
#  main)

```




SENTIMENT ANALYSIS


Now, let's take a look at the sentiments of the different groups' lyrics.

I'm using the Syuzhet package (with nrc lexicon) on the stemmed words and plotting on histograms to see the general shape of the distribution and boxplot for spread.

Though the sizes are vastly different, the general shapes of the distributions are similar and more or less concentrated at 0. The sentiments of other, according to the boxplot, do seem to have a wider spread, indicating more variation in emotional extremes.

```{r}

beatles.sentiments.nrc <- get_sentiment(beatles_lyrics$stemmedwords, method = "nrc", language = "english")
other.sentiments.nrc <- get_sentiment(other_lyrics$stemmedwords, method = "nrc", language = "english")

hist(beatles.sentiments.nrc, xlab = "Emotion index (nrc)", main = "Histogram of Beatles Sentiments", col = "blue")
hist(other.sentiments.nrc, xlab = "Emotion index (nrc)", main = "Histogram of Other Sentiments", col = "red")

```

```{r}
boxplot(beatles.sentiments.nrc, other.sentiments.nrc, horizontal = T, names = c("beatles", "other"), col = c("blue", "red"), xlab = "Emotion index (nrc)")
```


Using the 


```{r}
plot(beatles.sentiments.nrc, type = "h", main = "Beatles nrc Sentiment plot", ylab = "Emotional index")
plot(other.sentiments.nrc, type = "h",main = "Other nrc Sentiment plot", ylab = "Emotional index")
```


Even a piechart does not indicate a significant difference between the Beatles and other bands of its time, though they are slightly more positive than other bands.

```{r}
# emotion vector
emotions <- c("anger", "anticipation", "disgust", "fear", "joy", "sadness", "surprise", "trust", "negative", "positive")
# set sentiments to dataframe
beatles.sentiments.pie  <- data.frame(sentiments = colSums(get_nrc_sentiment(beatles_lyrics$stemmedwords)))
# this one may take a minute
other.sentiments.pie  <- data.frame(sentiments = colSums(get_nrc_sentiment(other_lyrics$stemmedwords)))


```


```{r}
# set emotions to dataframes of sentiments
beatles.sentiments.pie$emotions <- emotions
other.sentiments.pie$emotions <- emotions
# print out the sentiments for direct comparison
beatles.sentiments.pie
other.sentiments.pie

# piechart 
plot_ly() %>%
  add_pie(data = beatles.sentiments.pie, labels =  emotions, values = ~sentiments,
          name = "Beatles", domain = list(x = c(0, 0.4), y = c(0.4, 1))) %>%
  add_pie(data = other.sentiments.pie, labels =  emotions, values = ~sentiments,
          name = "Other", domain = list(x = c(0.6, 1), y = c(0.4, 1))) %>%
  layout(title = "Sentiments Pie Charts", showlegend = T,
         xaxis = list(showgrid = F, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = F, zeroline = FALSE, showticklabels = FALSE))


```




Conclusion:

- On average, when compared to other rock bands formed between 1960-1970, Beatles lyrics tend to be shorter than those of their counterparts.

- The most dominant words include love, baby, cry, and girl, which is not uncommon, though the Beatles do use cry relatively more than other bands.

- According to the piechart, the only difference between the Beatles and other bands is that the Beatles are slightly more positive and that other bands have a greater variance in emotional extremes.



